---
title: "Teaching an old dog new tricks? Learning rates, aging, and language change"
author: Ellis Cain
date: May 2, 2023
output:
  beamer_presentation:
    theme: "Frankfurt"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    fig_caption: False
classoption: "aspectratio=169"
bibliography: bibliography.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
library(here)
library(tidyverse)
library(wesanderson)
library(patchwork)
library(ggdist)
library(reshape2)
knitr::opts_knit$set(root.dir = here())
library(ggrepel)
library(jtools)
```


```{r data, include=FALSE}
## Setup ----
theme_set(theme_bw() + theme(text =  element_text(size = 14)))
`%nin%` = Negate(`%in%`)
```


# Background literature

## Collective patterns of language usage

<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->

- N-gram corpus analysis of Google Books corpus[^1]

- Diachronic changes in language usage and meaning[^2]

<!-- ::: -->
<!-- ::: {.column width="50%"} -->
\centering
![Examples of changes in lexical usage over time]("ngram_grammar_fig"){width=50%}
<!-- ::: -->
<!-- :::: -->

[^1]: [@michelQuantitativeAnalysisCulture2011]
[^2]: [@bybee2015language; @bynon1977historical]

## Collective patterns of language

<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->


*Emergence of collective-level trends from individual usage*

<!-- ::: -->
<!-- ::: {.column width="50%"} -->
\centering
![Examples of changes in lexical usage over time]("interaction_fig"){width=50%}
<!-- ::: -->
<!-- :::: -->

## Mechanisms of language acquisition

- Statistical learning[^3]

- Propose but verify (hypothesis testing)[^4]

- Structural inference[^5]

<!-- add more  -->

[^3]: [@smith2008infants; @yu2007rapid]
[^4]: [@trueswell2013propose]
[^5]: [@kim2019knowledge]

## Aging and learning

- Initially rely on associative / bottom-up learning

- Later shift to inferential / top-down learning

<!-- - *Make a point here* -->

# Formal model

## Overview

- Language change as interaction between individual and collective level dynamics

- Aim to explore how individual learning rates, aging, and group membership impact overall population-level patterns of language change

\pause

- Model of the usage and spread of a grammatical variant throughout a population

    - Past tense ending can be "-t" or "-ed", such as in "burnt" or "burned"
    
\centering
!["Burnt"]("ngram_burnt"){width=30%}
    
## Model assumptions[^6]

1. Language learning is based on imitating others, though this may change over the lifespan

    - E.g., individuals may learn quickly early on, but slow down as they age
    
\pause

2. There are variations in preference between individuals

    - E.g., some individuals learn more quickly than others
    
\pause

3. Language can be influenced by external factors

    - E.g., more willing to learn from in-group members
    
[^6]: [@beeksma2017probabilistic]

## Outcome patterns[^7]

1. **S-shaped curve in usage patterns**: Change happens slowly, then proceeds rapidly before slowing down again.

2. **Intra-speaker variation**: Change is gradual and there is a period of intra-speaker variation.

3. **Categorical norms**: With competition, speakers move toward categorically using just one of the competing variants.

4. **Multi-stability**: Language change can have multiple stable outcomes. *May result in dialect subgroups*

5. **Threshold problem**: Initially rare variants may manage to spread through entire speech communities.

[^7]: [@troutman2008social]


## Initialization: network

- Generates preferential attachment network

- Distributes grammar according to specified percentage of grammar 1

    - Two grammar variants, 0 or 1 (*burnt or burned*)

\centering
![]("example_network"){width=35%}

## Initialization: nodes

Represent language users

- State: node's current grammar preference, initialized as 0 or 1

- Age

    - Probablistic or deterministic

- Cohort: "Age group", *either 1 or 2*

- Gamma: learning rate of a given node

     - Probablistic, deterministic, or based on age

## Initialization: cohorts

- Number of cohorts (max 2) based on specified percentage

- Cohort ages can be specified

- Option for cohort-based grammar, such that the cohorts start with different percentages of grammar 1

- Willingness to listen to out-group members

## Dynamics

- Communication

    - Speaking (asynchronous)
    
    - Neighboring agents listen

- Aging

## Dynamics: speaking

- Nodes will generate an 'utterance', which is either 0 or 1 (*burnt or burned*)

- Nodes 'prefer' a discrete grammar

- Logistic curve is used when nodes produce an utterance

\centering
![]("troutman_fig6"){width=40%}

## Dynamics: listening

- Neighboring nodes will pick a grammar that will be used to interpret heard utterance

- If it matches the heard utterance

    - Update listener's state towards the heard state
    
    - Otherwise, it will update listener's state away from the heard state
    
    - Learning rate: *Gamma parameter modifies the step size*
    
- *Chance to ignore out-group*

## Dynamics: Aging

- Nodes age with each tick

- Gamma changes with age: either constant or decreasing with age

\pause

- Perseverance: how slowly gamma decays

    - Basic power law: $y = -0.05(x^{a})+0.005$
    
```{r gamma-plot, echo = F, fig.height = 2, fig.width = 4}
gamma.function = function(a, b, age, constant){
    return ( constant * (age^a) + b )
}

## Normal cost curve
a.exp = 1
initial.gamma = 0.005
front.constant = 0.05

normal.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# High cost 
a.exp = 2

high.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# Higher cost 
a.exp = 3

higher.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# Low cost 
a.exp = 0.5

low.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# Lower cost 
a.exp = 0.25

lower.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

df1 <- normal.cost |> 
    rbind(high.cost, higher.cost, low.cost, lower.cost) |> 
    mutate(type = "Increasing (q = 1)")

gamma.function = function(a, b, age, constant){
    return ( constant * (-age^a) + b )
}

## A as cost!!!

## Normal cost curve
a.exp = 1
initial.gamma = 0.055

normal.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# High cost 
a.exp = 2

high.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# Higher cost 
a.exp = 3

higher.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# Low cost 
a.exp = 0.5

low.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

# Lower cost 
a.exp = 0.25

lower.cost <- data.frame(age = seq(0,1,0.01)) |> 
    mutate(gamma = gamma.function(a.exp, initial.gamma, age, front.constant),
           a = a.exp, initial = initial.gamma, constant = front.constant)

df2 <- normal.cost |> 
    rbind(high.cost, higher.cost, low.cost, lower.cost) |> 
    mutate(type = "Decreasing (q = -1)")

df2 |> 
    # rbind(df2) |> 
    ggplot(aes(x = age, y = gamma, group = factor(a), color = factor(a)))+
    # geom_point()+
    geom_line(linewidth = 1)+
    theme_bw()+
    xlab("Proportional age (x)")+
    ylab("Gamma")+
    scale_color_brewer(name = "Perseverance (a)", palette = 5)

rm(high.cost, higher.cost, low.cost, normal.cost, lower.cost)
```

# Results

## Impact of learning rate

```{r iolr_data_import, echo = F}
gvdf_proc <- read_csv(here("paper","processed_data","gvdf_proc.csv"), show_col_types = F)
gvdf <- read_csv(here("paper","processed_data","gvdf_final-only.csv"), show_col_types = F)


```


```{r plot1.1, echo = F,  warning=F, message=F, fig.height=3, fig.width = 7}
ggplot(gvdf,
       aes(x = percent_grammar_1, y = mean_state_of_nodes,
           color = factor(initial_gamma),
           group = factor(initial_gamma)))+
  theme_bw()+
  xlab("Initial percentage of Grammar 1")+
  ylab("Mean state of nodes")+
  scale_color_brewer(name = "Initial gamma",# type = "qual",
                     palette = 5)+ 
  stat_summary(fun = mean, geom = "point")+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  geom_text(aes(x = 80, y = 0.8), label = "Grammar 1\nas majority", color = "black")+
  geom_text(aes(x = 15, y = 0.2), label = "Grammar 0\nas majority", color = "black")+
  labs(caption = "50 runs per configuration, 1000 steps total per run")
```

## Impact of learning rate

```{r plot1.2, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
ggplot(gvdf_proc, aes(x = initial_gamma, y = eq_percent,
                      group = factor(initial_gamma)))+
  xlab("Initial gamma")+
  ylab("Time to equilibrium (%)")+
  theme_bw()+
  stat_summary(fun = mean, geom = "bar", fill = "lightblue", color = "black")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.001)+
  labs(caption = "50 runs per configuration, 1000 steps total per run")
```


## Variation of learning rate amongst individual 

```{r volrai_data_import, echo = F}
lndf2_proc <- read_csv(here("paper","processed_data","lndf2_proc.csv"), show_col_types = F)
lndf2 <- read_csv(here("paper","processed_data","lndf2_final-only.csv"), show_col_types = F)
```

```{r plot2.1, echo = F,  warning=F, message=F, fig.height=3, fig.width = 7}
ggplot(lndf2,
       aes(x = percent_grammar_1, y = mean_state_of_nodes,
           color = factor(initial_gamma),
           group = factor(initial_gamma)))+
  theme_bw()+
  xlab("Initial percentage of Grammar 1")+
  ylab("Mean state of nodes")+
  scale_color_brewer(name = "Initial gamma",# type = "qual",
                     palette = 5)+ 
  stat_summary(fun = mean, geom = "point")+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  labs(caption = "Faceted by Gamma SD; 50 runs per configuration, 1000 steps total per run")+
  facet_wrap(~gamma_sd)
```

## Variation of learning rate amongst individual 

```{r plot2.2, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
ggplot(lndf2_proc, aes(x = initial_gamma, y = eq_percent,
                      group = factor(initial_gamma)))+
  xlab("Initial gamma")+
  ylab("Time to equilibrium (%)")+
  theme_bw()+
  stat_summary(fun = mean, geom = "bar", fill = "lightgreen", color = "black")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.001)+
  labs(caption = "50 runs per configuration, 1000 steps total per run")+
  facet_wrap(~gamma_sd)
```

## Checkpoint

- Increased gamma decrease time to equilibrium, but not (systematically) affect the final outcome


## Decrease in learning rate with age

```{r dilrwa_data_import, echo = F}
lndf3_proc <- read_csv(here("paper","processed_data","lndf3_proc.csv"), show_col_types = F)
det_age <- read_csv(here("paper","processed_data","lndf3_final-only.csv"), show_col_types = F)
```

```{r plot3.1, echo = F,  warning=F, message=F, fig.height=3, fig.width = 7}
det_age |> 
  ggplot(aes(x = percent_grammar_1, y = mean_state_of_nodes,
             color = factor(graph_group, levels = c("Deterministic", "Probabilistic (SD = 30)", "Probabilistic (SD = 50)", "Probabilistic (SD = 100)")),
             group = factor(graph_group, levels = c("Deterministic", "Probabilistic (SD = 30)", "Probabilistic (SD = 50)", "Probabilistic (SD = 100)"))))+
  theme_bw()+
  xlab("Initial percentage of Grammar 1")+
  ylab("Mean state of nodes")+
  scale_color_brewer(name = "Age type", type = "qual",
                     palette = 2)+
  stat_summary(fun = mean, geom = "point")+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  labs(caption = "50 runs per configuration, 1000 steps total per run")
```

## Decrease in learning rate with age

```{r plot3.2, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
lndf3_proc <- lndf3_proc |> 
  mutate(graph_group = if_else(deterministic_age == T,
                               "Deterministic",
                               if_else(deterministic_age == F & age_sd == 30,
                                       "Probabilistic (SD = 30)",
                                       if_else(deterministic_age == F & age_sd == 50,
                                               "Probabilistic (SD = 50)",
                                               if_else(deterministic_age == F & age_sd == 100,
                                                       "Probabilistic (SD = 100)",
                                                       "N/A")))))

ggplot(lndf3_proc, aes(x = factor(graph_group, levels = c("Deterministic", "Probabilistic (SD = 30)", "Probabilistic (SD = 50)", "Probabilistic (SD = 100)")),
                       y = eq_percent))+
  xlab("Deterministic age?")+
  ylab("Time to equilibrium (%)")+
  theme_bw()+
  stat_summary(fun = mean, geom = "bar", fill = "darkgreen", color = "black")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25)+
  labs(caption = "50 runs per configuration, 1000 steps total per run")
```

## Variation of speed of decrease (perseverance)

```{r vosodp_data_import, echo = F}
vgp_proc <- read_csv(here("paper","processed_data","vgp_proc.csv"), show_col_types = F)
vgp <- read_csv(here("paper","processed_data","vgp_final-only.csv"), show_col_types = F)
```

```{r plot4.1, echo = F,  warning=F, message=F, fig.height=3, fig.width = 7}
ggplot(vgp,
       aes(x = percent_grammar_1, y = mean_state_of_nodes,
           color = factor(perseverance),
           group = factor(perseverance)))+
  theme_bw()+
  xlab("Initial percentage of Grammar 1")+
  ylab("Mean state of nodes")+
  scale_color_brewer(name = "Perseverance",# type = "qual",
                     palette = 5)+ 
  stat_summary(fun = mean, geom = "point")+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  labs(caption = "50 runs per configuration, 1000 steps total per run")

```

## Variation of speed of decrease (perseverance)

```{r plot4.2, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
ggplot(vgp_proc, aes(x = factor(perseverance), y = eq_percent,
                       group = factor(perseverance)))+
  xlab("Perseverance of learning")+
  ylab("Time to equilibrium")+
  theme_bw()+
  stat_summary(fun = mean, geom = "bar", fill = "turquoise", color = "black")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25)+
  labs(caption = "50 runs per configuration, 1000 steps total per run")
```

## Checkpoint

- Gamma impacts time to equilibrium (TTE)

- No difference in TTE when perseverance is > 0.5

## Two age cohorts

```{r tac_data_import, echo = F}
tcg_proc <- read_csv(here("paper","processed_data","tcg_proc.csv"), show_col_types = F)
tcg_group_proc <- read_csv(here("paper","processed_data","tcg_group_proc.csv"), show_col_types = F)
tcg <- read_csv(here("paper","processed_data","tcg_final-only.csv"), show_col_types = F)
tcg_group <- read_csv(here("paper","processed_data","tcg_final-only-by-cohort.csv"), show_col_types = F)
```

```{r plot5.1, echo = F,  warning=F, message=F, fig.height=3, fig.width = 7}
ggplot(tcg |> 
         filter(step == 1000),
       aes(x = percent_grammar_1, y = mean_state_of_nodes,
           color = factor(percent_cohort2),
           group = factor(percent_cohort2)))+
  theme_bw()+
  xlab("Initial percentage of Grammar 1")+
  ylab("Mean state of nodes")+
  scale_color_brewer(name = "Cohort 2 percent",# type = "qual",
                     palette = 5)+ 
  stat_summary(fun = mean, geom = "point")+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  labs(caption = "Faceted by Cohort 2 age; 50 runs per configuration, 1000 steps total per run")+
  facet_grid(~cohort2_age)
```


## Two age cohorts

```{r plot5.2, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
ggplot(tcg_proc, aes(x = factor(percent_cohort2), y = eq_percent,
                     fill = factor(cohort2_age),
                     group = factor(percent_cohort2)))+
  xlab("Cohort 2 percentage")+
  ylab("Time to equilibrium")+
  theme_bw()+
  stat_summary(fun = mean, geom = "bar", color = "black")+
  scale_fill_brewer(name = "Group 2 age", type = "qual",
                     palette = 1)+ 
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25)+
  labs(caption = "Faceted by cohort 2 age; 50 runs per configuration, 1000 steps total per run")+
  theme(legend.position = "none")+
  facet_grid(~cohort2_age)
```

## Two age cohorts: group equilibria

```{r plot5.3, echo = F,  warning=F, message=F, fig.height=3, fig.width = 7}
tcg |> 
  filter(step == 1000) |> 
  select(c(percent_grammar_1, percent_cohort2, 
           cohort2_age, cohort1_state, cohort2_state)) |> 
  pivot_longer(cols = c(cohort1_state, cohort2_state), 
               values_to = "mean_state_of_nodes") |> 
  mutate(name = if_else(name == "cohort1_state", "Cohort 1", "Cohort 2")) |> 
  ggplot(aes(x = percent_grammar_1, y = mean_state_of_nodes,
           color = factor(percent_cohort2),
           group = factor(percent_cohort2)))+
  theme_bw()+
  xlab("Initial percentage of Grammar 1")+
  ylab("Mean state of nodes")+
  scale_color_brewer(name = "Group 2 percent",# type = "qual",
                     palette = 5)+ 
  stat_summary(fun = mean, geom = "point")+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  labs(caption = "50 runs per configuration, 1000 steps total per run")+
  facet_grid(name~cohort2_age)
```

## Two age cohorts: group equilibria

```{r plot5.4, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
ggplot(tcg_group_proc |> 
         mutate(type = if_else(type == "cohort1_equilibrium", "Cohort 1", "Cohort 2")) ,
       aes(x = factor(percent_cohort2), y = eq_percent,
                     fill = factor(cohort2_age),
                     group = factor(percent_cohort2)))+
  xlab("Cohort 2 percentage")+
  ylab("Time to equilibrium")+
  theme_bw()+
  stat_summary(fun = mean, geom = "bar", color = "black")+
  scale_fill_brewer(name = "Group 2 age", type = "qual",
                    palette = 1)+ 
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25)+
  labs(caption = "Faceted by cohort 2 age; 50 runs per configuration, 1000 steps total per run")+
  facet_grid(type~cohort2_age)+
  theme(legend.position = "none")
```

## Checkpoint

- Gamma impacts time to equilibrium (TTE)

- No difference in TTE when perseverance is > 0.5

- Older cohort slows down TTE

- As the age gap increases, the impact on TTE increases as well



## Cohort-based grammar: Only Cohort 2 has grammar 1

```{r dioc_data_import, echo = F}
vgg_g1v <- read_csv(here("paper","processed_data","vgg_types_group1_variation.csv"), show_col_types = F)
vgg_g2v <- read_csv(here("paper","processed_data","vgg_types_group2_variation.csv"), show_col_types = F)
```

```{r plot6.1, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 8}
ggplot(vgg_g2v, aes(x = type, y = final_state, group = type, fill = type, color = type))+
  xlab("Type")+
  ylab("Mean state")+
  theme_bw()+
  geom_violin()+
  # geom_density()+
  # geom_bar(stat = "bin", bins = 10, position = position_dodge())+
  labs(caption = "Gamma decreases at constant rate; Cohort 1 does not have grammar 1; 50% Cohort 2; 50 runs per configuration, 1000 steps total per run")+
  facet_grid(~c2_percent_grammar_1)+
  theme(legend.position = "none",
        axis.text.x = element_text(size = 6))
```



## Cohort-based grammar: Only Cohort 1 has grammar 1

```{r plot6.2, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 8}
ggplot(vgg_g1v, aes(x = type, y = final_state, group = type, fill = type, color = type))+
  xlab("Time to equilibrium")+
  theme_bw()+
  geom_violin()+
  # geom_density()+
  # geom_bar(stat = "bin", bins = 10, position = position_dodge())+
  labs(caption = "Gamma decreases at constant rate; Cohort 2 does not have grammar 1; 50% Cohort 2; 50 runs per configuration, 1000 steps total per run")+
  facet_grid(~c1_percent_grammar_1)+
  theme(legend.position = "none",
        axis.text.x = element_text(size = 6))

```


## Checkpoint

- Gamma impacts time to equilibrium (TTE)

- No difference in TTE when perseverance is > 0.5

- Older cohort slows down TTE, exacerbated by increased age gap

- Reservoir? When grammar varies only in the older cohort, their initial starting percentage will draw the younger cohort

- Innovator: when grammar varies only in the younger cohort, 
However, when cohort 2 does not use grammar 1, cohort 1 starts with increasingly higher percentages of using grammar 1, they themselves may end up with using grammar 1 more, but it does not drive the overall group usage.

## Cohort preference

```{r cp_data_import, echo = F}
pref_proc <- read_csv(here("paper","processed_data","preference_proc.csv"), show_col_types = F)
pref_types <- read_csv(here("paper","processed_data","preference_final_state.csv"), show_col_types = F)
```

```{r plot7.1, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
ggplot(pref_types, aes(x = percent_grammar_1, y = final_state,
             color = factor(chance_listen_to_outgroup),
             group = factor(chance_listen_to_outgroup)))+
  theme_bw()+
  xlab("Initial percent of grammar 1")+
  ylab("Final state")+
  scale_color_brewer(name = "Willingness to\nlisten to outgroup", type = "div",
                     palette = 2)+ 
  # scale_shape_manual(name = "type", values = c(14, 15, 16))+
  # geom_point(alpha = 0.5)+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  stat_summary(fun = mean, geom = "point", size = 2)+
  stat_summary(fun = mean, geom = "line")+
  # ggtitle("Different starting percentages")+
  labs(caption = "50 runs per configuration, 1000 steps total per run")+
  facet_grid(~factor(type))

```

## Cohort preference

```{r plot7.2, echo = F,  warning=F, message=F, fig.height=3.5, fig.width = 7}
ggplot(pref_proc, aes(x = percent_grammar_1, y = eq_percent,
                           color = factor(chance_listen_to_outgroup),
                           group = factor(chance_listen_to_outgroup)))+
  xlab("Percent grammar 1")+
  ylab("Time to equilibrium (%)")+
  theme_bw()+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 2)+
  stat_summary(fun = mean, geom = "point", size = 2)+
  stat_summary(fun = mean, geom = "line")+
  scale_color_brewer(name = "Willingness to\nlisten to outgroup", type = "div",
                    palette = 1)+ 
  labs(caption = "50 runs per configuration, 1000 steps total per run")

```

## Checkpoint

- Increased gamma speeds up when equilibrium is reached, but not the final outcome; modulated by perseverance

- Older cohort delays equilibrium

- "Innovator" and "Reservoir" groups

- Group preference will delay the equilibrium, and slightly impact equilibrium value

##

## References