---
title: "Teaching an old dog new tricks? Learning rates, aging, and language change"
author: Ellis Cain
date: April 2, 2023
output:
  beamer_presentation:
    theme: "Frankfurt"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    fig_caption: False
classoption: "aspectratio=169"
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(wesanderson)
library(patchwork)
library(ggdist)
library(reshape2)
knitr::opts_knit$set(root.dir = here())
library(ggrepel)
```


```{r data, include=FALSE}
## Setup ----
theme_set(theme_bw() + theme(text =  element_text(size = 14)))
`%nin%` = Negate(`%in%`)

## Read in the data ----
```


# Background literature

## Something to set it up

- Language change examples?

- Language acquisition?

## Language changes over time

- Google Books corpus (5.2 mill books) from 1800-2000[^1]

- Ran a "culturenomics" study through an n-gram corpus analysis

- Quantified trends in lexical usage, grammatical patterns, and social usage

[^1]: [@michelQuantitativeAnalysisCulture2011]

## Language changes over time

![Examples of changes in lexical usage over time]("ngram_fig")

## Verb Bias figures

- Verb bias figures

## *How* does language change over time

Types of lexical change[^2]:

- Metaphor: *head* or *face*

- Metonymy: *bureau* as cloth over a table

- Inference: *since* from temporal to causal

- Bleaching: *guys*

- Narrowing / generalization

[^2]: [@bybee2015language]

## Mechanisms of language acquisition

- Statistical learning

- Propose but verify (hypothesis testing)

- Structural inference

\pause

- Not exactly clear how these learning mechanisms may change over the lifespan, and how cognitive decline may impact executive functioning and learning.

## Learning rates

- MacMurray study: parallel learning, constant learning rate

- Blachstein study? Changes over the lifetime?

- Bryersbart study?

## Shift to modeling

- Previous models

- Main points



<!-- ## Age-related differences in lexico-semantic organization -->

<!-- - Word association data has been used to explore the organization of our lexico-semantic space[^3] -->

<!-- - Dubossarsky and colleagues[^4] analyzed lexical association networks -->

<!-- - Found that the **organization** changes across the lifespan -->

<!-- [^3]: [@dedeyneSmallWorldWords2019] -->
<!-- [^4]: [@dubossarskyQuantifyingStructureFree2017] -->

<!-- ## Age-related differences in lexico-semantic organization -->

<!-- U-shaped developmental trajectory: -->

<!-- Small and sparse $\rightarrow$ dense and well-connected $\rightarrow$ sparse with isolated nodes -->

<!-- ![Networks across the lifespan]("dubossarsky_fig") -->

<!-- ## Age-related differences in lexico-semantic processing -->

<!-- - EEG has also been used to explore the temporal aspect of lexico-semantic processing [^5] -->

<!-- - Previous study suggests that the neural signatures of predictive processing (increased frontal positivity) were reduced in older adults -->
<!-- \pause -->
<!-- - Aging may be related to changes in neural mechanisms of processing or the predictability among words/meanings -->

<!-- [^5]:[@federmeierAgerelatedIndividualDifferences2010; @wlotkoPredictNotPredict2012] -->

## Main research question

Given that adults learn throughout their life and update their linguistic representations[^8], it is especially relevant to understand:
\pause
**How do changes in meaning across time impact our understanding of those words?**

[^8]:[@brysbaert2014concreteness; @castro2021category; @ryskin2017verb]


## Model description:

- Overview

- Assumptions

## Agents/nodes

- Properties

## Initialization

- See chart

## Speaking

- Explanation

## Listening

- Explanation

## Modification: Gamma

- Explanation

- Modifications

## Modification: Aging

- Learning curve plot

## Analysis 1

## Analysis 2

## Conclusions

##

Thank you.

##

## References